\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2021-05-18 main.tex
finalise plots for paper

One of the most important applications of evolutionary game theory is the
evolution of cooperation. Why is it that some individuals choose to help others,
increasing their payoff, at the expense of decreasing one’s own? In evolutionary
game theory, individuals are not required to be rational, instead they adapt
strategies based on mutation and exploration. Strategies are more likely to
spread if they have a high fitness either because the individuals who adopt them
have more offspring, or because they are imitated more often. The fitness of a
strategy is not constant but depends on the composition of the population.
Individuals interact based on their strategies with other members of the
population and the payoffs they yield are translated into fitness.

It is commonly assumed that fitness is equivalent to the mean distribution of
types in the population. These payoffs assume that individuals can interact with
the entire population several times and remember each and every outcome. Thus,
they imply that individuals have a perfect memory. However, when they make
decisions in each turn they are assumed to have very limited memory. To be
precise, most of the works in the literature focus on naive subjects who can
only choose from a restricted set of strategies~\cite{Nowak1992tit}, or who do
not remember anything beyond the outcome of the very last round~\cite{Baek2016}.
Note that there are a few notable exceptions~\cite{Hauert1997, Stewart2016}.

The perfect memory assumption is not only unrealistic but it also creates this
curious inconsistency. This has led us to question how robust is our
understanding of cooperation. In this work, we explore whether direct
reciprocity can evolve if individuals only remember a minimum of social
information. Though we are not the first to question the assumptions of
estimating fitness~\cite{Roca2006}, we are the first to explore the effect of
payoff memory.

Initially, we consider two extreme scenarios. The first is the classical
scenario of the expected payoffs and the alternative scenario where individuals
update their strategies only based on the very last payoff they obtained. In the
later sections, we allow individuals more memory. More specifically, they can
remember up to two turns and up to two interactions. We present results on
several social dilemmas which include the prisoner’s dilemma and the donation
game, the snowdrift game, the stag hunt game and the harmony game.

The remainder of the paper is organized as follows. Section~\ref{section:model}
describes the model. Section~\ref{section:results} presents the results of the
simulations. Finally, section~\ref{section:conclusions} outlines the main
conclusions.

% Early results for the repeated prisoner’s dilemma and for the snowdrift game
% suggest that memory size can have a considerable effect on the evolutionary
% dynamics. As a rule of thumb, we observe that individuals with limited memory
% tend to adopt less generous strategies and they achieve less cooperation than in
% the classical scenario. In contrast, in the stag-hunt and the harmony game, the
% impact of memory is less striking. Here individuals with limited memory perform
% nearly as well as individuals with full memory.

@@

We consider a population of \(N\) players (\(N\) is even) where mutations are
sufficiently rare. Thus, at any point in time there are at most two
different strategies present in the population; a \textit{resident} strategy
and a \textit{mutant} strategy. We assume a pairwise process where
strategies spread because they are imitated more often.

@@

In this section we explore the case where the updating payoffs are based on the
last round payoff achieved against another member of the population. We compare
this to the expected payoffs. We assume that individuals interact in a donation
game  where each can cooperate by providing a benefit \(b\) to the other
player at their cost \(c\), with \(0 < c < b\). Thus, \(T=b, R=b-c, S=-c, P=0\).

@@

The figure suggests that when updating is based on expected payoffs, players
tend to be more generous and more cooperative. The $q$-values are higher on
average which suggests that individuals tend to be more generous when it comes
to cooperating after being at the receiving end of a defection. %This is a known result no? Nowak 1992
The average cooperation rate within the resident population is strictly higher
when expected payoffs are considered. The difference between the two methods for
both the values of \(b\) are statistically significant. This contrast becomes
more obvious for \(b=10\). More specifically the average cooperation drops from
97\% to 51\%. The residents of the population cooperated on average 97\% of the
times in one case and only 57\% in the other.

@@

We further explore the effect of the benefit,
Figure~\ref{fig:cooperation_rate_over_benefit}. The figure suggests that the
expected payoffs always overestimate cooperation. For the limited memory
payoffs, the cooperation rate remains unchanged at approximately 50\% once \(b =
5\). The highest cooperation rate that was achieved was 0.51 when  the last
round payoff is considered. In comparison, when the expected payoffs were used
the highest average cooperation rate was 0.97.

@@

Figure~\ref{fig:cooperation_rate_over_betas} illustrates results for various
runs of the evolutionary process where we vary the strength of selection. For
weak selection, \(\beta < 1\), the updating payoffs have no effect. The evolved
population and the average cooperating rate are the same for both approaches.
However for strong selection, it can be seen that the cooperating rate increases
with \(beta\), when expected payoffs are used. However, when limited memory
payoffs are used the cooperating rate decreases.

@@

By focusing on the donation game in the previous section, we have gained
insights into the possible effects of the updating payoffs, and how parameters
such as the benefit and the strength of selection might can exaggerate the
effect. However, by focusing on one specific game can narrow our knowledge of
the effects of updating payoffs on the different forms of possible human
interactions and social dilemmas. For this reason we explore not only the 
donation game but all the possible symmetric games. These are given by
Table~\ref{table:social_dilemmas}.

\begin{table}[!htbp]
  \begin{center}
    \resizebox{.6\textwidth}{!}{
    \renewcommand{\arraystretch}{1.5}
      \begin{tabular}{ccc}
  \toprule
  & \textbf{social dilemmas} & \textbf{payoffs' constrains} \\
  \midrule
(i) & stage hunt  & \(R > T > P > S\) \\
(ii) & snowdrift  & \(T > R > S > P\) \\
(iii) & harmony  & \(R > T > S > P\) \\
(iv) & prisoner dilemma  & \(T > R > P > S\) \\
(v)& donation game  & \(T > R > P > S\) \& \(T=b, R=b-c, S=-c, P=0\) \\
  \bottomrule
      \end{tabular}}
  \end{center}
  \caption{\textbf{Social dilemmas and payoffs' constrains}. The various social
  dilemmas we explore in this work. Results for cases (i) - (iv) are presented
  in section~\ref{section:2_by_2_games} and results for case (v) are presented
  in section~\ref{section:donation}.}
  \label{table:social_dilemmas}
\end{table}

Without loss of generality we set one of the entries of the payoff matrix to 0
and another one to 1. A common normalization is \(R=1\) and
\(P=0\)~\cite{Martinez2012, Roca2009}. We vary the temptation to defect \(T\)
and the risk of cooperating \(S\). We continue the discussion of the previous
section. We compare the expected payoffs to the last round payoffs over the
different social dilemmas. The results of the simulation runs are given in
Figures~\ref{fig:expected_two_by_two} and~\ref{fig:last_round_two_by_two}.

At the left up quadrat (\(S > 0\) and  \(T \leq 1\)) the individuals interact
in the harmony game at each game stage. The harmony game is a game without
tensions. The best play in the harmony game is to always cooperate. Indeed both
figures confirm this. The behaviour varies sightly at the case \(T = 1\). Now
the temptation is equal to the payoff of a cooperation and thus individuals now
tend to defect but only slightly. The cooperating rate still remains high. At
the bottom left up quadrat (\(S \leq 0\) and  \(T < 1\)) the individuals
interact in the stag hunt game at each game stage. In the stag hunt game
cooperation is still favoured. The best response of individuals is to cooperate
and this is again confirmed by both approaches.

The last two cases are those of the prisoner’s dilemma and of the snowdrift
game. At the top right quadrat we explore the snowdrift game. In the snowdrift
game individuals want to cooperate after receiving a defection and to defect
given that their opponent defected. This behavior is captured by the figures.
In the case of \(q\), cooperating after receiving a cooperation, we observe
a decrease as the suckers payoff decreases. This is to be expected. Finally,
in the right bottom quadrat individuals interact in a prisoner’s dilemma.
In the prisoner’s dilemma individuals tend to defect. In the case of direct
reciprocity cooperation can emerge, as we discussed in the previous section.
However, in the spectrum of all the different values of \(S\) and \(T\) used
here we observe this behaviour in only a handful of cases.

@@

We have discussed the expected behaviour for each social dilemma. We now want to
account the difference in the cooperation rate given the two updating payoffs.
The cooperation rates for each of the case of the social dilemmas we have used
are given in Figure~\ref{fig:cooperation_rates_two_by_two_part_one}.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=.45\textwidth]{static/expected_two_by_two_games_cooperation.pdf}
  \includegraphics[width=.45\textwidth]{static/stochastic_two_by_two_games_cooperation.pdf}
  \caption{\textbf{Cooperation rates over various social dilemmas.} ({\bf A}) If
  players update based on their expected payoffs. ({\bf B}) If
  players update based on their last round payoffs.}\label{fig:cooperation_rates_two_by_two_part_one}
\end{figure}

>>>>

One of the most important applications of evolutionary game theory is the
evolution of cooperation. Why is it that some individuals choose to help others,
increasing their payoff, at the expense of decreasing one’s own? In evolutionary
game theory, individuals are not required to be rational, instead they adapt
strategies based on mutation and exploration. Strategies are more likely to
spread if they have a high fitness either because the individuals who adopt them
have more offspring, or because they are imitated more often~\cite{Wu2015}. The
fitness of a strategy is not constant but depends on the composition of the
population. Individuals interact based on their strategies with other members of
the population and the payoffs they yield are translated into fitness.

It is commonly assumed that individuals compute their fitness after interacting
with a representative sample of the population, and remembering all the
interactions they participated in~\cite{Nowak2006}. Thus, they imply that
individuals have a perfect memory. However, when modelling how individuals make
decisions in each turn they are assumed to have very limited memory. To be
precise, most of the works in the literature focus on naive subjects who can
only choose from a restricted set of strategies~\cite{Nowak1992tit}, or who do
not remember anything beyond the outcome of the very last round~\cite{Baek2016}.
Note that there are a few notable exceptions~\cite{Hauert1997, Stewart2016}.
Thus, the perfect memory assumption is not only unrealistic but it also creates
this curious inconsistency.

This has led us to question the robustness of our understanding of cooperation,
In this work we explore whether cooperation can evolve if individuals compute
their fitness based on a minimum of social information. Though we are not the
first to question the assumptions of estimating fitness~\cite{Roca2006}, we are
the first to explore the effect of payoff memory.

We first consider two extreme scenarios, the classical scenario of the expected
payoffs and the alternative scenario where individuals update their strategies
only based on the very last payoff they obtained. We observe that individuals
with limited memory tend to adopt less generous strategies and they achieve less
cooperation than in the classical scenario. We obtain similar results when we
consider that individuals update their strategies based on more social
information. More specifically, up to the last two payoffs they obtained when
interacting with up to two different members of the population.

The remainder of the paper is organized as follows. In
section~\ref{section:model} we describe the model. In
section~\ref{section:results} we presents the results of the simulations.
Finally, section~\ref{section:conclusions} we outline the main conclusions.

@@

We consider a population of \(N\) players~\footnote{The terms ``player'' and
``individual'' are used interchangeably here.} (\(N\) is even) where mutations are
sufficiently rare. Thus, at any point in time there are at most two different
strategies present in the population; a \textit{resident} strategy and a
\textit{mutant} strategy. We assume a pairwise process where strategies spread
because they are imitated more often.

@@

In this section we explore the case where the updating payoffs are based on the
last round payoff achieved against another member of the population. We compare
this to the classical scenario of the expected payoffs. We assume that each pair
of players interacts in a donation game. The donation game is a special case of
the prisoner's dilemma. Each player can choose to cooperate by providing a
benefit \(b\) to the other player at their cost \(c\), with \(0 < c < b\).
Thus, the payoffs of a one round interact are \(T=b, R=b-c, S=-c, P=0\).

@@

The figure suggests that when updating is based on expected payoffs, players
tend to be more generous and more cooperative. The $q$-values of the resident
strategies are on average higher in the case of the expected payoffs. The
players will occasionally forgive a defection more often if their fitness
depends on interacting with every member of the population. On the other hand,
when social interactions are limited they are less forgiving. The average
cooperation rate for each simulation is calculated as the the average
cooperation rate within the resident population. In the case of the expected
payoffs, regardless the value of benefit, the average cooperation rate is
strictly higher than that of the last round payoffs. The difference based the
two methods is statistically significant, and the dissimilarity is even more
obvious for a high value of \(b\). More specifically, the average cooperation of
resident strategies drops from 97\% to 57\%.

@@

We further explore the effect of the benefit.
Figure~\ref{fig:cooperation_rate_over_benefit} suggests that expected payoffs
always yield a higher cooperation rate. We observe that the cooperation rate
increases as the value of the benefit gets higher. In comparison, for the
limited memory payoffs, the cooperation rate remains unchanged, at approximately
50\%.

@@

We also investigate the effect of the strength of selection.
Figure~\ref{fig:cooperation_rate_over_betas} illustrates results for various
runs of the evolutionary process. For weak selection, \(\beta < 1\), we observe
that the two methods yield the same result, however, as \(\beta\) increases
there is variation in the evolving populations. In the case of expected payoffs
the resident populations become more and more cooperative as \(\beta\)
increases, whereas in the case of limited memory payoffs, the resident
populations become less cooperative.

@@

In the previous subsection we gained insights into the effects of the updating
payoffs, and into how parameters such as the benefit and the strength of
selection can intensify them. We investigated these effects by using the
donation game. In order to broaden our understanding of the updating payoffs on
different forms of possible human interactions, we extend our approach to other
\(2 \times 2\) symmetric games. More specifically, we apply our analysis to the
four different classes of games given by Table~\ref{table:social_dilemmas}. We
compare the results of the evolutionary process described in
section~\ref{section:model} when the updating payoffs are based on the expected
payoffs, and on the last round payoffs for all the four possible classes of
games. The results of the simulations are given in
Figures~\ref{fig:expected_two_by_two} and~\ref{fig:last_round_two_by_two}
respectively.

\begin{table}[!htbp]
  \begin{center}
    \resizebox{.45\textwidth}{!}{
    \renewcommand{\arraystretch}{1.5}
      \begin{tabular}{ccc}
  \toprule
  & \textbf{social dilemmas} & \textbf{preference ordering} \\
  \midrule
(iii) & harmony  & \(R > T > S > P\) \\
(i) & stag hunt  & \(R > T > P > S\) \\
(iv) & prisoner dilemma  & \(T > R > P > S\) \\
(ii) & snowdrift  & \(T > R > S > P\) \\
  \bottomrule
      \end{tabular}}
  \end{center}
\caption{\textbf{Social dilemmas and payoffs' constrains}. The various social
dilemmas we explore in this work. Results for cases (i) - (iv) are presented
in section~\ref{section:2_by_2_games} and results for case (v) are presented
in section~\ref{section:donation}.}
\label{table:social_dilemmas}
\end{table}

In Figure~\ref{fig:expected_two_by_two} each sub figure represents a run of
evolutionary process for a different set of values the payoffs of an individual,
\(\mathcal{U}\). Without loss of generality we set \(R=1\) and
\(P=0\)~\cite{Martinez2012, Roca2009}, and we vary the values of the temptation
\(T\) (across the \(x\)-axis) and of the risk of cooperating \(S\) (across the
\(y\)-axis). There are four classes of games as presented in
Table~\ref{table:social_dilemmas}. Starting at the upper left corner and proceeding
clockwise correspond to the harmony game, the snowdrift, the prisoner's dilemma
and the stag hunt.

The harmony game represents a situation without tensions where it is in the best
interest of both players to cooperate. This is confirmed by our simulations
Figure~\ref{fig:expected_two_by_two}. In the several experiments of the harmony
game the resident population applies a strategy for which \(p \approx 1\) and
\(q \approx 1\).

The snowdrift game describes a situation similar to the prisoner's dilemma where
cooperation results in a benefit to the opposing player but entailing a cost to
the cooperator. Compared to the prisoner's dilemma, individuals obtain immediate
direct benefits from the cooperative acts. The story of the snowdrift game
usually goes as follow: two drivers are trapped on either side of a snowdrift
and have the options of staying in the car or removing the snowdrift. Letting
the opponent do all the work is the best option but if the other player stays in
the car it is better to shove. In the end cooperation leads to the cooperated
drive to get home, even though they had to endure the work, thus  \(S > P\). In
the snowdrift game more cooperation emerges compared to the prisoner's dilemma.
Defection is never a resident strategy and for the shame values of temptation
the overall \(q-values\) are higher. It can be seen that the \(p-values\) are
lower as individuals have free room to defect after receiving a cooperation.

The last class of games corresponds to the stag hunt game. In the stag hunt game
both players benefit for mutual cooperation and defection, however, \(R > P\) and
thus the resident strategies with the highest fitness are the cooperative ones.
The results are similar to the cases of the harmony game.

@@

The results for the last round payoffs are fairly similar with the patterns
begging the same. It can been seen from Figure~\ref{fig:last_round_two_by_two}
that the results are move variation. We have discussed the expected behaviour
for each social dilemma. We now want to account the difference in the
cooperation rate given the two updating payoffs. The cooperation rates for each
of the case of the social dilemmas we have used are given in
Figure~\ref{fig:cooperation_rates_two_by_two_part_one}.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=.4\textwidth]{static/expected_two_by_two_games_cooperation.pdf}
  \includegraphics[width=.4\textwidth]{static/stochastic_two_by_two_games_cooperation.pdf}
  \caption{\textbf{Cooperation rates over various social dilemmas.} ({\bf A}) If
  players update based on their expected payoffs. ({\bf B}) If
  players update based on their last round payoffs.}\label{fig:cooperation_rates_two_by_two_part_one}
\end{figure}
